\textsc{Rils}-\textsc{Rols} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-11 paper_resources/doc/paper.tex
Revisions: Abstract and Conclusions

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known, prominent problem of Symbolic regression, intensively studied within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science, and finding models for band gaps of NaCl-type compounds
		in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to   determine the best--fitting   coefficients in considered models (solutions);  ($iii$) a novel fitness function is utilized which combines an accuracy score, error score, and the weighted size of the model into one, more compact score. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} algorithm to 14 other competitors from the literature considering different levels of noise in the input data.  Our method obtains new best exact percentages on both benchmark sets compared to other literature approaches. Moreover, a new unbiased randomly generated set of problem instances are introduced where the scalability of our method is thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

In this paper we dealt with solving the well--known Symbolic regression (SR)  problem. SR problem is a generalization of more interdisciplinary appearing  linear and logistic regression problems. %SR problem have been intensively studied in the literature over the last 15 years, where a few dozens of algorithms were constructed, mainly based on Genetic programming and on the application of different kinds of neural networks. 
We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. Our approach is built upon   the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the ordinary least square method to efficiently determine internal coefficients that fit the candidate solution,  thus in this way  the search process is focused on considering more promising solutions. Another key aspect is the utilization of a carefully constructed fitness function, which appears as  a combination of three important scores: RMSE, $R^2$  scores and the size of model (solution).  Last but not least, a carefully constructed local search is applied systematically exploring solutions obtained as 1--perturbations of the considered solution.  1--perturbations of a solution are based on applying the six per-node change operations on the solution's tree representation.

\textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from literature on the two ground truth benchmark sets from literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed efficiently of our method which produced the best average ranking results in terms of exact percentage when compared to all other competitors on the two ground-truth benchmark sets, in presence of all three different levels of noise in the input data. In all cases, new best exact percentages are obtained by means of \textsc{Rils-Rols} method. In some cases, these differences are  statistically significant in favor of our \textsc{Rils-Rols} method. Moreover, scalability and robustness of this method is  checked and confirmed on the unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of model, the number of variables in exact models, and average runtimes.

In the future work, one could think of constructing a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics to further boost quality of the  obtained results. For example, replacing ILS scheme of \textsc{Rils-Rils} with  a more general Variable neighborhood search (VNS) scheme is a reasonable option to try out. VNS   utilizes more stronger and systematic system of solutions diversification  and thus   escaping from local optima could be much easier.  Additionally, as a proof of efficiency and benefits of our algorithm, this tool could  be applied to solving problems from various problem domains out of the areas of computer science, for example from physics or chemistry, to help for setting up hypotheses or to obtain valuable insights on   experimental evaluations.

>>>>

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known, prominent  Symbolic regression problem, intensively studied within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping to establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science,   finding models for band gaps of NaCl-type compounds
		in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used in the search to make it efficient with  determining the best--fitting   coefficients in considered solutions (models);  ($iii$) a novel fitness function is utilized which combines the three known measures: $R^2$ score, RMSE score, and the size of the model which is weighted by a parameter. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} algorithm to 14 other competitors from the literature where different levels of the Gaussian white noise in the input data are utilized.  Our method obtains new best exact percentages on both benchmark sets compared to other literature approaches regardless of varying the level of nose. Moreover, a new unbiased randomly generated set of problem instances are introduced where the scalability of our method is thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

In this paper, we dealt with solving the well--known Symbolic regression (SR)  problem. SR problem is a generalization of more interdisciplinary known problems of linear and logistic regression.    
We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. Our approach is built upon the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the ordinary least square method to efficiently determine internal coefficients that are the best fitting to the candidate solution. Thus, the search process is primarily focused on considering already promising solutions in the continuous part of the search space rather than applying enhanced heuristics to search for those coefficients. Another key aspect is the utilization of a carefully constructed fitness function in the search, which appeared as a combination of three important scores: RMSE, $R^2$  scores and the size of the model (solution).  Last but not least, a carefully constructed local search is applied systematically exploring solutions obtained as 1--perturbations of the considered solution.  1--perturbations of a solution are based on applying the six per-node change operations on the solution's tree representation.

\textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from the literature on the two ground truth benchmark sets from literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed the efficiency of our method which produced the best average ranking results in terms of exact percentage when compared to all other competitors on the two ground-truth benchmarks in the presence of all three different levels of noise in the input data. In all cases, new best exact percentages are obtained using \textsc{Rils-Rols} method.  
Moreover, the scalability and robustness of this method are checked and confirmed on the unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of the model, the number of variables included in exact models, and obtained average runtimes.

In future work, one could think of constructing a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics to further boost the quality of the obtained results. For example, replacing ILS scheme of \textsc{Rils-Rils} with a more general Variable neighbourhood search (VNS) scheme is a reasonable option to give it a try. VNS   utilizes a more stronger and systematic system of solutions diversification and thus escaping from local optima could be much easier.  Additionally, as proof of the efficiency and benefits of our algorithm, this tool could be applied to solving problems from various problem domains out of the area of computer science, for example from physics or chemistry, to help for setting up hypotheses or to obtain valuable insights on experimental evaluations.
